---
title: "fs_weather_tables"
author: Natalie Novoseltseva
format: html
---

Measuring Climate Change Effects through Water Quality

Field Station Weather Data EDA

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#libraries
library(tidyverse)
library(dplyr)
library(lubridate)
library(gt)
library(ggplot2)
library(stringr)
```

## Data

```{r}
#field station meteorological data 2001-2026

#creates a data folder
dir.create("data/field_station_weather", recursive = TRUE, showWarnings = FALSE) 

#using tempfile() to create a tempoarry storage
zip <- tempfile(fileext = ".zip")

#downlod the file from the url
download.file("http://cdmo.baruch.sc.edu/aqs/output/854334.zip", zip, mode = "wb")

unzip(zip, exdir = "data/field_station_weather")

#read in the files for all the years as a one data file hudfsmet-p
files <- list.files(
  "data/field_station_weather",
  pattern = "^hudfsmet[0-9]{4}\\.csv$", # ^ -starts a string
                                       # hudfsmet - the exact name in the file (constant)
                                       # any digits from 0-9 might follow
                                       # {4} - we are looking for 4 digits pattern
                                       # \\. - there should be a dot
                                       # csv - it should end with "csv"
                                       # end a string
  full.names = TRUE
)

fs_weather_01_26_raw <- do.call(
  rbind,
  lapply(files, function(f) {
    df <- read.csv(f, fileEncoding = "Latin1")
    df$year <- as.integer(gsub(".*([0-9]{4})\\.csv$", "\\1", f))
    df
  })
)
```

## Data Cleaning

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Removing empty columns and other irrelant columns
fs_weather_01_26 <- fs_weather_01_26_raw |>
  remove_empty("cols")|>
  select(-StationCode,
        -isSWMP, #historical and provisional Plus columns indicate the status of QAQC
        -Historical, 
        -ProvisionalPlus,
  )

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

#creating a vector with all the flags
flags_weather <- c(
  "F_Record",
  "F_ATemp",
  "F_RH",
  "F_BP",
  "F_WSpd",
  "F_MaxWSpd",
  "F_Wdir",
  "F_SDWDir",
  "F_TotPAR",
  "F_TotPrcp",
  "F_TotSoRad")

flagged_data <- "<-3>|<-2>|<1>"

fs_weather_01_26 <- fs_weather_01_26 |>
  mutate(ATemp = if_else(str_detect(F_ATemp, flagged_data), NA_real_, ATemp),
         RH = if_else(str_detect(F_RH, flagged_data), NA_real_, RH),
         BP = if_else(str_detect(F_BP, flagged_data), NA_real_, BP),
         WSpd = if_else(str_detect(F_WSpd, flagged_data), NA_real_, WSpd),
         MaxWSpd = if_else(str_detect(F_MaxWSpd, flagged_data), NA_real_, MaxWSpd),
         Wdir = if_else(str_detect(F_Wdir, flagged_data), NA_real_, Wdir),
         SDWDir = if_else(str_detect(F_SDWDir, flagged_data), NA_real_, SDWDir),
         TotPAR = if_else(str_detect(F_TotPAR, flagged_data), NA_real_, TotPAR),
         TotPrcp = if_else(str_detect(F_TotPrcp, flagged_data), NA_real_, TotPrcp)
  )

#removing flag columns 
fs_weather_01_26 <- fs_weather_01_26 |> 
  select(-all_of(flags_weather))

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

#separate the data and time column
fs_weather_01_26 <- fs_weather_01_26 %>%
  mutate(
    DatetimeStamp = as.POSIXct(DatetimeStamp, format = "%m/%d/%Y %H:%M"), #converts the variable into standard POSIXct format
    date = as.Date(DatetimeStamp),
    time = format(DatetimeStamp, "%H:%M:%S")
  )

#create a month column
fs_weather_01_26 <- fs_weather_01_26 %>%
  mutate(
    month = month(DatetimeStamp)          
  )

#create a season column
fs_weather_01_26 <- fs_weather_01_26 %>%
  mutate(
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% 3:5        ~ "Spring",
      month %in% 6:8        ~ "Summer",
      month %in% 9:11       ~ "Fall"
    )
  )
```

## Tables and averages

Number of observations

```{r}
#extract observations with missing season values
fs_weather_01_26_NAs <- fs_weather_01_26 %>%
  filter(is.na(season))

#count observations by year and season and display as a formatted table
fs_weather_01_26 %>%
  count(year, season) %>%
  gt(groupname_col = "season") %>%
  tab_header(
    title = "Observation Coverage by Year and Season"
  ) %>%
  cols_label(
    year = "Year",
    n = "Number of Observations"
  )

```

Summary statistics (all years)

```{r}
#create a summary table from the dataset
weather_summary <- fs_weather_01_26 %>%
#calculate summary statistics (mean and standard deviation) for each variable
  summarise(
    Mean_ATemp    = mean(ATemp, na.rm = TRUE),
    SD_ATemp      = sd(ATemp, na.rm = TRUE),
    Mean_RH       = mean(RH, na.rm = TRUE),
    SD_RH         = sd(RH, na.rm = TRUE),
    Mean_BP       = mean(BP, na.rm = TRUE),
    SD_BP         = sd(BP, na.rm = TRUE),
    Mean_WSpd     = mean(WSpd, na.rm = TRUE),
    SD_WSpd       = sd(WSpd, na.rm = TRUE),
    Mean_MaxWSpd  = mean(MaxWSpd, na.rm = TRUE),
    SD_MaxWSpd    = sd(MaxWSpd, na.rm = TRUE),
    Mean_TotPAR   = mean(TotPAR, na.rm = TRUE),
    SD_TotPAR     = sd(TotPAR, na.rm = TRUE),
    Mean_TotPrcp  = mean(TotPrcp, na.rm = TRUE),
    SD_TotPrcp    = sd(TotPrcp, na.rm = TRUE)
  ) %>%
  #convert the wide summary output into a long format
  tidyr::pivot_longer(
    cols = everything(),
    names_to = c("Statistic", "Variable"),
    names_sep = "_",
    values_to = "Value"
  ) %>%
  #convert back to wide format so Mean and SD become columns
  tidyr::pivot_wider(
  tidyr::pivot_wider(
    names_from = Statistic,
    values_from = Value
  ) %>%
  #replace abbreviated variable names with descriptive labels
  mutate(
    Variable = recode(
      Variable,
      ATemp   = "Air Temperature (°C)",
      RH      = "Relative Humidity (%)",
      BP      = "Barometric Pressure (hPa)",
      WSpd    = "Wind Speed (m/s)",
      MaxWSpd = "Maximum Wind Speed (m/s)",
      TotPAR  = "Total PAR (µmol/m²)",
      TotPrcp = "Total Precipitation (mm)"
    )
  )
#create a formatted table using the gt package
weather_summary %>%
  gt() %>%
  tab_header(
    title = "Summary Statistics for Weather Variables",
    subtitle = "Hudson River Field Station Weather Data (2001–2026)"
  ) %>%
  cols_label(
    Variable = "Variable",
    Mean = "Mean",
    SD = "Standard Deviation"
  ) %>%
  fmt_number(
    columns = c(Mean, SD),
    decimals = 2
  )
```

Averages by season

```{r}
#group weather data by season
fs_weather_01_26 %>%
  group_by(season) %>%
#calculate seasonal averages
  summarise(
    avg_air_temp = mean(ATemp, na.rm = TRUE),
    avg_humidity = mean(RH, na.rm = TRUE),
    avg_wind_spd = mean(WSpd, na.rm = TRUE),
    avg_precip   = mean(TotPrcp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
#convert to a formatted table
  gt() %>%
  tab_header(
    title = "Average Weather Conditions by Season",
    subtitle = "Hudson River Field Station (2001–2026)"
  ) %>%
#rename columns
  cols_label(
    season        = "Season",
    avg_air_temp  = "Air Temperature (°C)",
    avg_humidity  = "Relative Humidity (%)",
    avg_wind_spd  = "Wind Speed (m/s)",
    avg_precip    = "Total Precipitation (mm)"
  ) %>%
  fmt_number(
    decimals = 2
  )
```

Average temperatures by year

```{r}
avg_temp_by_year <- fs_weather_01_26 %>%
  group_by(year) %>%
  summarise(
    avg_air_temp = mean(ATemp, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )

avg_temp_by_year %>%
  gt() %>%
  tab_header(
    title = "Average Air Temperature by Year",
    subtitle = "Hudson River Field Station Weather Data (2001–2026)"
  ) %>%
  fmt_number(
    columns = avg_air_temp,
    decimals = 2
  ) %>%
  cols_label(
    year = "Year",
    avg_air_temp = "Average Air Temperature (°C)",
    n_obs = "Observations"
  )
```

Average temperature by year and season (one table)

```{r}
#calculate average air temperature for each year
avg_air_temp_by_year_season <- fs_weather_01_26 %>%
  group_by(season, year) %>%
  summarise(
    avg_air_temp = mean(ATemp, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(season))

#display results as a formatted table
avg_air_temp_by_year_season %>%
  gt(groupname_col = "season") %>%
  tab_header(
    title = "Average Air Temperature by Year and Season",
    subtitle = "Hudson River Field Station Weather Data (2001–2026)"
  ) %>%
  fmt_number(
    columns = avg_air_temp,
    decimals = 2
  ) %>%
  cols_label(
    year = "Year",
    avg_air_temp = "Average Air Temperature (°C)",
    n_obs = "Observations"
  )
```

Seasonal extremes

```{r}
#calculate seasonal air temperature extremes
fs_weather_01_26 %>%
  group_by(season) %>%
  summarise(
    min_air_temp = min(ATemp, na.rm = TRUE),  #lowest observed air temperature
    max_air_temp = max(ATemp, na.rm = TRUE),  #highest observed air temperature
    p95_air_temp = quantile(ATemp, 0.95, na.rm = TRUE), #95th percentile (excludes extreme outliers)
    .groups = "drop"
  ) %>%
#display results as a formatted table
  gt() %>%
  tab_header(
    title = "Seasonal Air Temperature Extremes",
    subtitle = "Hudson River Field Station Weather Data (2001–2026)"
  ) %>%
  fmt_number(
    decimals = 2
  ) %>%
  cols_label(
    season = "Season",
    min_air_temp = "Minimum Air Temperature (°C)",
    max_air_temp = "Maximum Air Temperature (°C)",
    p95_air_temp = "95th Percentile Air Temperature (°C)"
  )
```

Changes in the magnitude of temperature changes

```{r}
#calculate average air temperature for each year
avg_air_temp_by_year <- fs_weather_01_26 %>%
  group_by(year) %>% #group data by year
  summarise(
    avg_air_temp = mean(ATemp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    delta_overall = avg_air_temp - mean(avg_air_temp, na.rm = TRUE) # difference from overall average
  )

#display results as a formatted table
avg_air_temp_by_year %>%
  gt() %>%
  tab_header(
    title = "Average Air Temperature by Year",
    subtitle = "Deviation from Overall Mean (2001–2026)"
  ) %>%
  fmt_number(
    columns = c(avg_air_temp, delta_overall),
    decimals = 2
  ) %>%
  cols_label(
    year = "Year",
    avg_air_temp = "Average Air Temperature (°C)",
    delta_overall = "Difference from Overall Average (°C)"
  )
```

Plot of the magnitude of temperature changes

```{r}
#calculate average air temperature for each year
avg_air_temp_by_year <- fs_weather_01_26 %>%
  group_by(year) %>%
  summarise(
    avg_air_temp = mean(ATemp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    delta_overall = avg_air_temp - mean(avg_air_temp, na.rm = TRUE) #difference from overall average
  )

#calculate the overall mean temperature across all years
overall_mean <- mean(avg_air_temp_by_year$avg_air_temp)

#plot yearly average air temperature
ggplot(avg_air_temp_by_year, aes(x = factor(year), y = avg_air_temp, fill = delta_overall > 0)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = overall_mean, linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "Average Air Temperature by Year",
    subtitle = "Deviation from Overall Mean (2001–2026)",
    x = "Year",
    y = "Average Air Temperature (°C)"
  ) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "lightgray")) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Plot of the magnitude of temperature changes by season

```{r}
#calculate average air temperature by year and season
avg_air_temp_by_year_season <- fs_weather_01_26 %>%
  filter(!is.na(season)) %>%  # exclude missing seasons
  group_by(season, year) %>%
  summarise(
    avg_air_temp = mean(ATemp, na.rm = TRUE),
    .groups = "drop"
  )

#calculate overall seasonal averages (one per season)
seasonal_mean <- avg_air_temp_by_year_season %>%
  group_by(season) %>%
  summarise(season_mean = mean(avg_air_temp, na.rm = TRUE), .groups = "drop")

#merge seasonal mean back to the yearly data
avg_air_temp_by_year_season <- avg_air_temp_by_year_season %>%
  left_join(seasonal_mean, by = "season") %>%
  mutate(
    delta_overall = avg_air_temp - season_mean
  )

#plot
ggplot(avg_air_temp_by_year_season, aes(x = factor(year), y = avg_air_temp, fill = delta_overall > 0)) +
  geom_col(show.legend = FALSE) +
  geom_hline(aes(yintercept = season_mean), color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~season, scales = "free_y") +
  labs(
    title = "Average Air Temperature by Year and Season",
    subtitle = "Bars show yearly averages; red dashed line = seasonal mean (2001–2026)",
    x = "Year",
    y = "Average Air Temperature (°C)"
  ) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "lightgray")) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
