---
title: "Exploratory Data Analysis"
author: Ilianna Ditren
---

# Project Set Up

```{r, message=FALSE, warning=FALSE}

#Downloading libraries
library(readr)
library(tidyverse)
library(dplyr)
library(lubridate)
library(janitor)
library(rlang)
library(glue)

```

```{r, message=FALSE, warning=FALSE}

# Importing stony creek data
stony_creek <- read_csv("data/stony_creek.csv")

```

```{r, message=FALSE, warning=FALSE}

wq_clean_names <- c(
  Temp = "Water Temperature (C)",
  SpCond = "Specific Conductivity (mS/cm)",
  Sal = "Salinity (ppt)",
  DO_Pct = "Dissolved Oxygen (% Saturation)",
  DO_mgl = "Dissolved Oxygen (mg/L)",
  Depth = "Sonde Depth (m)",
  pH = "pH",
  ChlFluor = "Chlorophyll Fluorescence (ug/L)", # include foot note
  Turb = "Turbidity (NTU)",
  ATemp = "Average Air Temperature (C)",
  RH = "Average Relative Humidity (% Saturation)",
  BP = "Average Barometric Pressure (mb)",
  WSpd = "Average Wind Speed (m/s)",
  MaxWSpd = "Max Wind Speed (m/s)",
  MaxWSpdT = "Time of Max Wind Speed Measured (hh:mm)",
  Wdir = "Average Wind Direction", # include foot note
  SDWDir = "Wind Direction (sd)",
  TotPAR = "Total Photosynthetically Active Radiation",
  TotPrcp = "Total Precipitation (mm)",
  year = "Year",
  time = "Time",
  month = "Month",
  season = "Season",
  date = "Date",
  week_n = "Week")
```

# Climate Change Effects of Water Quality (20-years change)

## Exploratory Data Analysis

The first step in our exploratory data was to assess the descriptive statistics of each variable in the dataset. Our first presentation showed tables and line graphs of the mean values by season by year. The feedback we received revealed that linegraphs were not the best strategy to present the data because once we calculated the mean, the data stopped being continuous, instead, a scatterplot was a better visualization. In addition, we decided to removed the flagged data recommended by the client to have better results with the data. That process was done in the processing document

The following code creates a function to average the data by year and place it in a scatterplot. This function can also be found in the package called wqD375. The package was created in attempts to support data analysis of this project. The second chunk of code is a for loop to make efficient the creation of 5 plots (one for each of the seasons and one for all seasons) for each of the variables. 
The purpose of this tasks was to investigate whether or not there were major changes in water quality across the 20 years. 

```{r, echo = FALSE, message = FALSE}

scatterplot_mean <- function(df, 
                             var1, 
                             var2, 
                             season, 
                             aes_color = NULL, 
                             ymin = NULL, 
                             ymax = NULL) {
  
  # creating
  var1_name <- as_name(ensym(var1))
  var2_name <- as_name(ensym(var2))

# accessing the variable name from the wq_clean_names object and assigning it to the x and y labels
  x_label <- wq_clean_names[[var1_name]]
  y_label <- wq_clean_names[[var2_name]]

  df |>
    filter(!is.na({{var2}}) & season %in% {{season}}) |>
    group_by(year, season) |>
    summarize(mean = mean({{var2}})) |>
    ggplot() +
    geom_point(aes(x = {{var1}}, y = mean, color = {{aes_color}})) +
    labs(
      title = glue("Mean {season} {y_label} by Year"),
      x = x_label,
      y = y_label) +
    theme_minimal() +
    ylim({{ymin}}, {{ymax}}) +
    scale_color_viridis_d()
}

seasons = c("Spring", "Summer", "Fall", "Winter")
list_seasons <- list("Spring", "Summer", "Fall", "Winter", seasons)

```

```{r, warning = FALSE, message = FALSE}

for (lst in list_seasons) {
  print(scatterplot_mean(stony_creek, year, Temp, lst, season, 0, 23))
}

for (lst in list_seasons) {
  print(scatterplot_mean(stony_creek, year, SpCond, lst, season, 0, 1))
}

for (lst in list_seasons) {
  print(scatterplot_mean(stony_creek, year, Sal, lst, season, 0, 1))
}

for (lst in list_seasons) {
  print(scatterplot_mean(stony_creek, year, DO_Pct, lst, season, 0, 155))
}

for (lst in list_seasons) {
  print(scatterplot_mean(stony_creek, year, DO_mgl, lst, season, 0, 23))
}

for (lst in list_seasons) {
  print(scatterplot_mean(stony_creek, year, Depth, lst, season, 0, 1))
}

for (lst in list_seasons) {
  print(scatterplot_mean(stony_creek, year, pH, lst, season, 1, 14))
}

for (lst in list_seasons) {
  print(scatterplot_mean(stony_creek, year, Turb, lst, season, 0, 50))
}

for (lst in list_seasons) {
  print(scatterplot_mean(stony_creek, year, ChlFluor, lst, season, 0, 5))
}

```
