---
title: "joined_dataset_processing"
format: html
---

This document contains the processing steps for the main datasets. Initially, water quality data collected from Stony Creek and weather data collected from Field Station were explored individually. In the following steps, we combine the two datasets in order to perform multivariate exploratory data analysis.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#libraries
library(tidyverse)
library(dplyr)
library(lubridate)
library(stringr)
library(janitor)
library(reshape2)
```

Data
```{r}
# reading main datasets
water <- read_csv("Spring2026/data/stony_creek.csv")
weather <- read_csv("Spring2026/data/fs_weather_01_26.csv")
```

Exploring the structure of the data

```{r}
str(weather)
str(water)

names(weather)
names(water)
```

```{r}
#rename the column in the water dataset to easier join the datasets
weather <- weather %>%
  rename(DateTimeStamp = DatetimeStamp)
```

Joining the water quality and weather datasets
```{r}
#join water and weather data with inner join
combined <- inner_join(water, weather, by = "DateTimeStamp")
```

```{r}
#get rid of the duplicate columns and 
combined <- combined %>%
  select(-year.y, -date.y, -time.y, -month.y, -season.y) %>%
  rename(
    year = year.x,
    date = date.x,
    time = time.x,
    month = month.x,
    season = season.x
  )
```

```{r}
#the code to write out the dataset
write_csv(combined, "Spring2026/data/combined_dataset/joined_dataset.csv")
```

The combined dataset contained variables measured every 15-30 minutes. In this section, the data will be summarized to contain daily averages only. 

```{r}

#creating a vector with all numerical variable that need mean computed and extracting a name list

raw_combined <- combined |>
  mutate(season_num = case_when(
    season == "Winter" ~ 1,
    season == "Spring" ~ 2,
    season == "Summer" ~ 3,
    season == "Fall" ~ 4,
  ))

# computing the daily mean values for all variables in the combined_names object
daily_combined <- raw_combined |>
  group_by(date) |> 
  summarize(across(
    where(is.numeric), ~ mean(.x, na.rm = TRUE)),.groups = "drop")

#writing new csv
write_csv(daily_combined, "Spring2026/data/combined_dataset/daily_combined.csv")
   
```

In this chunk, the function `week()` is doing all the work of grouping the dates and assigning them a value from 1 to 52 based on which number of week the grouped dates fall in.

```{r}

# weekly averages of all variables in the combined dataset
weekly_combined <- combined |>
  mutate(
    week_n = week(DateTimeStamp), 
    year = year(DateTimeStamp)) |>
  group_by(week_n, year) |>
  summarize(across(
    where(is.numeric), ~ mean(.x, na.rm = TRUE)),.groups = "drop") |>
  select(-month, -Frequency)

#writing new csv
write_csv(weekly_combined, "Spring2026/data/combined_dataset/weekly_combined.csv")
             
```
