---
title: "This is my title"
author: "Ilianna Ditren, Natalie Novoseltseva, Joy Terentiev"
format: revealjs

---
```{r}
#libraries
library(tidyverse)
library(readr)
library(dplyr)
library(lubridate)
library(janitor)
library(stringr)
library(reshape2)
library(gt)
library(here)


```

```{r}
#data
stony_creek <- read_csv("data/stony_creek.csv")
combined <- read_csv("data/combined.csv")

```


## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Introduction
</div>


## Project Background

- Focus area: Stony Creek, NY
- Initial analysis covered 2020–2021 data
- Original project analyzed:
  - Temperature
  - pH
  - Salinity

## Our Contributions

- Expanded analysis to all available years (2002-2021)
- Combined water quality data with weather data
- Explored:
  - Long-term trends
  - Weather–water relationships
  - Seasonal patterns
  - Anomalies and possible data errors
  
## Research Question

**How did water quality indicators change over time, and were these changes associated with weather conditions?**

- Exploratory analysis only
- First step toward studying climate-related environmental change

## What the Report Includes

- Data cleaning and preparation
- Exploratory analysis
- Visualizations
- Seasonal analysis

## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Data and Methods
</div>

## Data Sources

- Stony Creek water quality data (2002–2021)
- Hudson Field Station weather data (2001–2026)

- Data source: Centralized Data Management Office (CDMO)
  https://cdmo.baruch.sc.edu

## Data Cleaning

- Removed empty rows and columns 

- Separated data and time column and added a year, month and season column

- Removed flagged data and then removed columns

## Final Dataset

- `combined`: Creates one dataset by joining water quality data from Stony Creek and meteorological data from Hudson Field Station.

## Main Variables

| Column | Description | 
|---------------|-------------------------------------------|---------------|
| Turb | Turbidity measured in units referred to as FNU/NTU, depending upon sensor.|
| TotPrcp | total precipitation measured in millimeters (mm) |
| SpCond | Specific conductivity measured in milli-Siemens per centimeter (mS/cm) |


## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Initial Exploratory Analysis
</div>


## Summary Statistics

```{r}
summary_tbl <- stony_creek %>%
  summarise(
    Temp_mean = mean(Temp, na.rm = TRUE),
    Temp_sd   = sd(Temp, na.rm = TRUE),
    Temp_min  = min(Temp, na.rm = TRUE),
    Temp_max  = max(Temp, na.rm = TRUE),

    DO_mean   = mean(DO_mgl, na.rm = TRUE),
    DO_sd     = sd(DO_mgl, na.rm = TRUE),
    DO_min    = min(DO_mgl, na.rm = TRUE),
    DO_max    = max(DO_mgl, na.rm = TRUE),

    pH_mean   = mean(pH, na.rm = TRUE),
    pH_sd     = sd(pH, na.rm = TRUE),
    pH_min    = min(pH, na.rm = TRUE),
    pH_max    = max(pH, na.rm = TRUE),

    Sal_mean  = mean(Sal, na.rm = TRUE),
    Sal_sd    = sd(Sal, na.rm = TRUE),
    Sal_min   = min(Sal, na.rm = TRUE),
    Sal_max   = max(Sal, na.rm = TRUE),

    Turb_mean = mean(Turb, na.rm = TRUE),
    Turb_sd   = sd(Turb, na.rm = TRUE),
    Turb_min  = min(Turb, na.rm = TRUE),
    Turb_max  = max(Turb, na.rm = TRUE),

    ChlFluor_mean = mean(ChlFluor, na.rm = TRUE),
    ChlFluor_sd   = sd(ChlFluor, na.rm = TRUE),
    ChlFluor_min  = min(ChlFluor, na.rm = TRUE),
    ChlFluor_max  = max(ChlFluor, na.rm = TRUE),

    SpCond_mean = mean(SpCond, na.rm = TRUE),
    SpCond_sd   = sd(SpCond, na.rm = TRUE),
    SpCond_min  = min(SpCond, na.rm = TRUE),
    SpCond_max  = max(SpCond, na.rm = TRUE)
  ) %>%
  pivot_longer(
    everything(),
    names_to = c("Variable", "Statistic"),
    names_sep = "_",
    values_to = "Value"
  ) %>%
  mutate(
    Variable = recode(
      Variable,
      Temp = "Temperature",
      DO = "Dissolved Oxygen",
      pH = "pH",
      Sal = "Salinity",
      Turb = "Turbidity",
      ChlFluor = "Chlorophyll",
      SpCond = "Conductivity"
    )
  ) %>%
  pivot_wider(
    names_from = Variable,
    values_from = Value
  ) 
```

- The table summarizes the key statistics for the main water quality variables in the Stony Creek dataset from 2002 to 2021.

```{r}
summary_tbl %>%
  gt() %>%
  fmt_number(decimals = 2) %>%
  tab_options(
    table.width = pct(100),
    table.font.size = px(22),
    heading.title.font.size = px(28),
    data_row.padding = px(6)
  ) %>% 
  tab_source_note(
    source_note = "Data Source: Centralized Data Management Office (CDMO), https://cdmo.baruch.sc.edu"
  )
```


## 

- Examining seasonal means shows that water temperature and temperature-dependent variables such as dissolved oxygen, turbidity, and conductivity vary significantly across seasons. Among these, it is notable that the summer–fall and winter–spring values are relatively similar for some variables.

```{r}
stony_creek %>%
  filter(!is.na(season)) %>%
  group_by(season) %>%
  summarise(
    Temperature        = mean(Temp, na.rm = TRUE),
    `Dissolved Oxygen` = mean(DO_mgl, na.rm = TRUE),
    pH                 = mean(pH, na.rm = TRUE),
    Salinity           = mean(Sal, na.rm = TRUE),
    Turbidity          = mean(Turb, na.rm = TRUE),
    Chlorophyll        = mean(ChlFluor, na.rm = TRUE),
    Conductivity       = mean(SpCond, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(season = str_to_title(season)) %>%
  gt() %>%
  tab_header(
    title = "Average Water Quality Measures by Season"
  ) %>%
  fmt_number(decimals = 2) %>%
  tab_options(
    table.width = pct(100),
    table.font.size = px(22),
    heading.title.font.size = px(28),
    data_row.padding = px(6)
  ) %>% 
  tab_source_note(
    source_note = "Data Source: Centralized Data Management Office (CDMO), https://cdmo.baruch.sc.edu"
  )
```

## {.stretch}

```{r, fig.width=14, fig.height=7}
avg_temp_by_year <- stony_creek %>%
  group_by(year) %>%
  summarise(
    avg_temp = mean(Temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    delta_overall = avg_temp - mean(avg_temp, na.rm = TRUE)
  )

overall_mean <- mean(avg_temp_by_year$avg_temp)

ggplot(avg_temp_by_year, aes(x = factor(year), y = avg_temp, fill = delta_overall > 0)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = overall_mean, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Average Water Temperature by Year",
    subtitle = "Yearly averages relative to the overall mean (2001–2021)",
    x = "Year",
    y = "Average Water Temperature (°C)"
  ) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "lightgray")) +
  theme_minimal(base_size = 18) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- The plot shows yearly average water temperature from 2001–2021, the red dashed line represents the mean temperature, blue bars indicate above-average years, gray below-average years.

##

```{r}
avg_temp_by_year_season <- stony_creek %>%
  filter(!is.na(season)) %>%
  group_by(year, season) %>%
  summarise(
    avg_temp = mean(Temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(season) %>%
  mutate(
    delta_overall = avg_temp - mean(avg_temp, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(avg_temp_by_year_season,
       aes(x = factor(year), y = avg_temp, fill = delta_overall > 0)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ season, ncol = 2) +
  geom_hline(
    data = avg_temp_by_year_season %>%
      group_by(season) %>%
      summarise(overall_mean = mean(avg_temp)),
    aes(yintercept = overall_mean),
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Average Water Temperature by Year and Season",
    subtitle = "Yearly seasonal averages relative to seasonal means (2001–2021)",
    x = "Year",
    y = "Water Temperature (°C)"
  ) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "lightgray")) +
  theme_minimal(base_size = 16) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(2, "lines")  
  )
```

## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Joined Dataset Analysis
</div>

##

```{r}

#select + rename variables
cor_vars <- combined %>%
  select(
    AirTemperature      = ATemp,
    RelativeHumidity    = RH,
    BarometricPressure  = BP,
    WindSpeed           = WSpd,
    MaxWindSpeed        = MaxWSpd,
    WindDirection       = Wdir,
    WindDirectionSD     = SDWDir,
    PAR                 = TotPAR,
    Precipitation       = TotPrcp,

    WaterTemperature    = Temp,
    DissolvedOxygen     = DO_mgl,
    OxygenPercent       = DO_Pct,
    Salinity            = Sal,
    Conductivity        = SpCond,
    pH                  = pH,
    Turbidity           = Turb,
    Chlorophyll         = ChlFluor,
    Depth               = Depth
  ) %>%
  select(where(~ sum(!is.na(.)) > 0))

#correlation matrix
cor_matrix <- cor(cor_vars, use = "pairwise.complete.obs")

# keep only lower triangle
cor_matrix[upper.tri(cor_matrix)] <- NA

#convert to long format
cor_long <- reshape2::melt(cor_matrix, na.rm = TRUE)
names(cor_long) <- c("Variable1", "Variable2", "Correlation")

#plot

ggplot(cor_long, aes(x = Variable1, y = Variable2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), size = 3) +
  scale_fill_gradient2(
    low = "#2b83ba",
    mid = "white",
    high = "#d7191c",
    midpoint = 0,
    name = "Correlation (r)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Relationships Between Weather and Water Quality Variables",
    x = NULL,
    y = NULL
  )
```




## Precipitation vs Specific Conductivity {.stretch}

```{r, fig.width=12, fig.height=7}
combined %>%
  filter(!is.na(TotPrcp), !is.na(SpCond), !is.na(season)) %>%
  ggplot(aes(x = TotPrcp, y = SpCond)) +
  geom_point(alpha = 0.4, color = "#2C7BB6", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#D7191C", linewidth = 1) +
  facet_wrap(~ season) +
  labs(
    title = "Precipitation and Specific Conductivity by Season",
    x = "Total Precipitation (mm)",
    y = "Specific Conductivity (mS/cm)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Lagged Effects of Precipitation on Conductivity {.stretch}

```{r, fig.width=12, fig.height=7}
#filter the data to have equal breaks
combined_1 <- combined %>%
  filter(year>2006)

#create lags 
combined_1 <- combined_1 %>%
  arrange(DateTimeStamp) %>%
  mutate(
    prcp_1hr  = lag(TotPrcp, 4),
    prcp_3hr  = lag(TotPrcp, 12),
    prcp_6hr  = lag(TotPrcp, 24),
    prcp_12hr = lag(TotPrcp, 48),
    prcp_24hr = lag(TotPrcp, 96)
  ) %>%
  select(SpCond, Turb, Temp, prcp_1hr, prcp_3hr, prcp_6hr, prcp_12hr, prcp_24hr) %>%
  pivot_longer(cols = starts_with("prcp_"),
               names_to = "lag",
               values_to = "precip")

#renaming columns to prepare for plots and releveling to keep order
combined_1 <- combined_1 %>%
  mutate(lag = case_when(lag == "prcp_1hr" ~ "1hr after rain",
                         lag == "prcp_3hr" ~ "3hr after rain",
                         lag == "prcp_6hr" ~ "6hr after rain",
                         lag == "prcp_12hr" ~ "12hr after rain",
                         lag == "prcp_24hr" ~ "24hr after rain"))

combined_1 <- combined_1 %>%
  mutate(lag = fct_relevel(lag,
                           "1hr after rain",
                           "3hr after rain",
                           "6hr after rain",
                           "12hr after rain",
                           "24hr after rain"))
                           

combined_1 %>%
  filter(!is.na(SpCond), !is.na(precip)) %>%
  ggplot(aes(x = precip, y = SpCond)) +
  geom_point(alpha = 0.25, color = "#2C7BB6") +
  geom_smooth(method = "lm", color = "#D7191C", se = TRUE) +
  facet_wrap(~ lag, scales = "free_x") +
  labs(
    title = "Lagged Precipitation vs Specific Conductivity",
    x = "Precipitation (lagged)",
    y = "Specific Conductivity (mS/cm)"
  ) +
  theme_minimal(base_size = 16) +
  theme(strip.text = element_text(face = "bold"))

```

## Precipitation vs Turbidity {.stretch}

```{r, fig.width=12, fig.height=7}

combined |>
  filter(!is.na(TotPrcp), !is.na(Turb), !is.na(season)) %>%
  ggplot(aes(x = TotPrcp, y = Turb)) +
  geom_point(alpha = 0.4, color = "#2C7BB6", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#D7191C", linewidth = 1) +
  facet_wrap(~ season) +
  labs(
    title = "Precipitation and Turbidity by Season",
    x = "Total Precipitation (mm)",
    y = "Water Turbidity"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Lagged Effects of Precipitation on Water Turbidity {.stretch}

```{r, fig.width=12, fig.height=7}
#filter the data to have equal breaks
combined_1 %>%
  filter(!is.na(Turb), !is.na(precip)) %>%
  ggplot(aes(x = precip, y = Turb)) +
  geom_point(alpha = 0.25, color = "#2C7BB6") +
  geom_smooth(method = "lm", color = "#D7191C", se = TRUE) +
  facet_wrap(~ lag, scales = "free_x") +
  labs(
    title = "Lagged Precipitation vs Water Turbidity",
    x = "Precipitation (lagged)",
    y = "Water Turbidity"
  ) +
  theme_minimal(base_size = 16) +
  theme(strip.text = element_text(face = "bold"))
```

## Precipitation vs Water Temperature {.stretch}

```{r, fig.width=12, fig.height=7}

combined |>
  filter(!is.na(TotPrcp), !is.na(Temp), !is.na(season)) %>%
  ggplot(aes(x = TotPrcp, y = Turb)) +
  geom_point(alpha = 0.4, color = "#2C7BB6", size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#D7191C", linewidth = 1) +
  facet_wrap(~ season) +
  labs(
    title = "Precipitation and Water Temperature by Season",
    x = "Total Precipitation (mm)",
    y = "Water Temperature (C)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Lagged Effects of Precipitation on Water Temperature {.stretch}

```{r, fig.width=12, fig.height=7}
#filter the data to have equal breaks
combined_1 %>%
  filter(!is.na(Temp), !is.na(precip)) %>%
  ggplot(aes(x = precip, y = Temp)) +
  geom_point(alpha = 0.25, color = "#2C7BB6") +
  geom_smooth(method = "lm", color = "#D7191C", se = TRUE) +
  facet_wrap(~ lag, scales = "free_x") +
  labs(
    title = "Lagged Precipitation vs Water Turbidity",
    x = "Precipitation (lagged)",
    y = "Water Temperature"
  ) +
  theme_minimal(base_size = 16) +
  theme(strip.text = element_text(face = "bold"))
```


## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Conclusion
</div>

## 

- Several data quality issues were identified, including missing values and possible sensor errors.
- Some plots suggest higher temperatures in recent years, especially during summer, but no clear long-term warming trend was confirmed.
- Interesting relationships were observed, particularly between precipitation and water temperature.
- Improved approaches were identified, including averaging data by day or week and analyzing seasons separately.

## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Next Steps
</div>

##

- Extend the analysis beyond exploratory methods by applying statistical modeling and time-series analysis.
- Analyze extreme weather events separately to better understand their environmental impacts and associated water quality changes.
- Improve data quality assessment by identifying sensor anomalies and patterns of missing data.
- Develop predictive models to better understand how water quality responds to precipitation and seasonal weather conditions.




