---
title: "This is my title"
author: "Ilianna Ditren, Natalie Novoseltseva, Mikhal"
format: revealjs

---
```{r}
#libraries
library(tidyverse)
library(readr)
library(dplyr)
library(lubridate)
library(janitor)
library(stringr)
library(reshape2)
library(gt)
library(here)


```

```{r}
#data
stony_creek <- read_csv("data/stony_creek.csv")

```


## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Introduction
</div>


## Project Background

- Focus area: Stony Creek, NY
- Initial analysis covered 2020–2021 data
- Original project analyzed:
  - Temperature
  - pH
  - Salinity

## Our Contributions

- Expanded analysis to all available years (2002-2021)
- Combined water quality data with weather data
- Explored:
  - Long-term trends
  - Weather–water relationships
  - Seasonal patterns
  - Anomalies and possible data errors
  
## Research Question

**How did water quality indicators change over time, and were these changes associated with weather conditions?**

- Exploratory analysis only
- First step toward studying climate-related environmental change

## What the Report Includes

- Data cleaning and preparation
- Exploratory analysis
- Visualizations
- Seasonal analysis

## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Data and Methods
</div>

## Data Sources

- Stony Creek water quality data (2002–2021)
- Hudson Field Station weather data (2001–2026)

- Data source: Centralized Data Management Office (CDMO)
  https://cdmo.baruch.sc.edu


## Data Cleaning

-  

## Final Dataset

-   


## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Initial Exploratory Analysis
</div>


## Summary Statistics

```{r}
summary_tbl <- stony_creek %>%
  summarise(
    Temp_mean = mean(Temp, na.rm = TRUE),
    Temp_sd   = sd(Temp, na.rm = TRUE),
    Temp_min  = min(Temp, na.rm = TRUE),
    Temp_max  = max(Temp, na.rm = TRUE),

    DO_mean   = mean(DO_mgl, na.rm = TRUE),
    DO_sd     = sd(DO_mgl, na.rm = TRUE),
    DO_min    = min(DO_mgl, na.rm = TRUE),
    DO_max    = max(DO_mgl, na.rm = TRUE),

    pH_mean   = mean(pH, na.rm = TRUE),
    pH_sd     = sd(pH, na.rm = TRUE),
    pH_min    = min(pH, na.rm = TRUE),
    pH_max    = max(pH, na.rm = TRUE),

    Sal_mean  = mean(Sal, na.rm = TRUE),
    Sal_sd    = sd(Sal, na.rm = TRUE),
    Sal_min   = min(Sal, na.rm = TRUE),
    Sal_max   = max(Sal, na.rm = TRUE),

    Turb_mean = mean(Turb, na.rm = TRUE),
    Turb_sd   = sd(Turb, na.rm = TRUE),
    Turb_min  = min(Turb, na.rm = TRUE),
    Turb_max  = max(Turb, na.rm = TRUE),

    ChlFluor_mean = mean(ChlFluor, na.rm = TRUE),
    ChlFluor_sd   = sd(ChlFluor, na.rm = TRUE),
    ChlFluor_min  = min(ChlFluor, na.rm = TRUE),
    ChlFluor_max  = max(ChlFluor, na.rm = TRUE),

    SpCond_mean = mean(SpCond, na.rm = TRUE),
    SpCond_sd   = sd(SpCond, na.rm = TRUE),
    SpCond_min  = min(SpCond, na.rm = TRUE),
    SpCond_max  = max(SpCond, na.rm = TRUE)
  ) %>%
  pivot_longer(
    everything(),
    names_to = c("Variable", "Statistic"),
    names_sep = "_",
    values_to = "Value"
  ) %>%
  mutate(
    Variable = recode(
      Variable,
      Temp = "Temperature",
      DO = "Dissolved Oxygen",
      pH = "pH",
      Sal = "Salinity",
      Turb = "Turbidity",
      ChlFluor = "Chlorophyll",
      SpCond = "Conductivity"
    )
  ) %>%
  pivot_wider(
    names_from = Variable,
    values_from = Value
  )
```

## Summary Statistics

- The table summarizes the key statistics for the main water quality variables in the Stony Creek dataset from 2002 to 2021.

```{r}
summary_tbl %>%
  gt() %>%
  fmt_number(decimals = 2) %>%
  tab_options(
    table.width = pct(100),
    table.font.size = px(22),
    heading.title.font.size = px(28),
    data_row.padding = px(6)
  ) %>% 
  tab_source_note(
    source_note = "Data Source: Centralized Data Management Office (CDMO), https://cdmo.baruch.sc.edu"
  )
```






## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Joined Dataset Analysis
</div>


## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Seasonal Analysis
</div>


## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Discussion
</div>

## {.center background-image="pictures/stony_creek_picture.jpeg"}

<div style="
  background: rgba(255,255,255,0.85);
  padding: 20px 50px;
  border-radius: 6px;
  font-size: 2.5em;
  font-weight: 600;
  display: inline-block;
">
Next Steps
</div>


## Summary Statistics

-   

## Summary Stat Boxplots - 2020

```{r, result = FALSE, eval=FALSE}
library(tidyverse)
library(readr)
dataset_2020 <- read_csv("Environmental_Data/data/hudscwq2020.csv")
dataset_2021 <- read_csv("Environmental_Data/data/hudscwq2021.csv")

dataset_2020 <- dataset_2020 %>%
  separate(DateTimeStamp, into = c("date", "time"), sep = " ")%>%
  mutate(date_clean = mdy(str_trim(date)))

dataset_2021 <- dataset_2021 %>%
  separate(DateTimeStamp, into = c("date", "time"), sep = " ")%>%
  mutate(date_clean = mdy(str_trim(date)))

dataset_2020 <- dataset_2020 %>%
  mutate(month = month(date_clean))

dataset_2021 <- dataset_2021 %>%
  mutate(month = month(date_clean))

dataset_2020 <- dataset_2020 %>%
  mutate(date_clean = as.Date(date_clean))

dataset_2021 <- dataset_2021 %>%
  mutate(date_clean = as.Date(date_clean))

dataset_2020 <- dataset_2020 %>%
  mutate(
    month = as.numeric(month), #make sure its considered a numeric because maybe not all months are being read correctly since there is section of gray of unlabelled season data in the two plots.
    season = case_when(
      month %in% 3:5 ~ "Spring",
      month %in% 6:8 ~ "Summer",
      month %in% 9:11 ~ "Fall",
      month %in% c(12, 1, 2) ~ "Winter",
      TRUE ~ NA_character_ #clarifies that only values of 12,1 and 2 should go into winter to avoid mislabelling
    )
  )

#copy for 2021
dataset_2021 <- dataset_2021 %>%
  mutate(
    season = case_when(
      month %in% 3:5 ~ "Spring",
      month %in% 6:8 ~ "Summer",
      month %in% 9:11 ~ "Fall",
      month %in% c(12, 1, 2) ~ "Winter",
      TRUE ~ NA_character_ #clarifies that only values of 12,1 and 2 should go into winter to avoid mislabelling
    )
  )

no_winter_2020 <- dataset_2020 %>%
  filter(season != "Winter")

no_winter_2021 <- dataset_2021 %>%
  filter(season != "Winter")


```

```{r, result = FALSE, eval=FALSE}
no_winter_2020 %>%
  pivot_longer(
    cols = c(Temp, Sal, Turb, ChlFluor, Level, Depth, pH, SpCond, DO_Pct, DO_mgl),
    names_to = "variable",
    values_to = "value") %>%
  ggplot(aes(x = season, y = value, fill = season))+
  geom_boxplot()+
  facet_wrap(~ variable, scales = "free_y")+
  scale_fill_manual(values = c(
    "Fall" = "orange",
    "Spring" = "springgreen4",
    "Summer" = "red"
  ))+
  labs(title = "Distribution of Variables by Season 2020",
       x = "Season",
       y = "Value")
```

## Summary Boxplots - 2021

```{r, result = FALSE, eval=FALSE}
no_winter_2021 %>%
  pivot_longer(
    cols = c(Temp, Sal, Turb, ChlFluor, Level, Depth, pH, SpCond, DO_Pct, DO_mgl),
    names_to = "variable",
    values_to = "value") %>%
  ggplot(aes(x = season, y = value, fill = season))+
  geom_boxplot()+
  facet_wrap(~ variable, scales = "free_y")+
  scale_fill_manual(values = c(
    "Fall" = "orange",
    "Spring" = "springgreen4",
    "Summer" = "red"
  ))+
  labs(title = "Distribution of Variables by Season 2021",
       x = "Season",
       y = "Value")
```

## Temperature Exploration

-   Initial plots using geom_line() and geom_point() reveled predictable seasonal variations
-   Summer peaks and Fall declines: consistent for both years
-   For monthly comparison, July and August consistently warm months (though 2021 still had missing data for some of this period)
-   Winter data sparse and so it was discluded from future visualizations

## Salinity Exploration

-   I created bar plots and faceted by month in order to visualize frequencies of salinity levels on a granular level
-   This variable was focused on due to the nature of the variable
-   It is a raw measurement and therefore I later rounded it into bins.
-   Monthly visualizations showed interesting differences in salinity levels, particularly the spread of them. For example, in November, 2020 salinity levels concentrate almost entirely at 0.2. In contrast, in September, levels were much more spread between 0.1 and 0.4.

## Salinity Continued

-   I also explored other ways to visualize Salinity data for both years
-   I later took out the monthly facets and observed proportionality seasonally by year.
-   These plots also show the difference in spread between 2020 and 2021, though 2021 does have more missing data which may partially influence this.

## Salinity Plots Proportionally - 2020

```{r, result = FALSE, eval=FALSE}
ggplot(data = no_winter_2020, aes(x = Sal, fill = season))+ geom_bar(position = "fill")+
  ylab("Proportion")+
  scale_fill_manual(values = c(
  "Fall"   = "orange",
  "Spring" = "springgreen4",
  "Summer" = "red"
))+
  scale_x_continuous(limits = c(0, 0.5))+
  labs(title = "Salinity by Season in 2020",
       x = "Salinity",
       y = "Proportion")
```

## Salinity Proportionally - 2021

```{r, result = FALSE, eval=FALSE}
ggplot(data = no_winter_2021, aes(x = Sal, fill = season))+ geom_bar(position = "fill")+
  ylab("Proportion")+
  scale_fill_manual(values = c(
  "Fall"   = "orange",
  "Spring" = "springgreen4",
  "Summer" = "red"
))+
  scale_x_continuous(limits = c(0, 0.5))+
  labs(title = "Salinity by Season in 2021",
       x = "Salinity",
       y = "Proportion")
```

## Temperature and pH Visual Exploration

-   I created boxplots to look at seasonal temperature distributions specificially
-   I also used density plots to look at a more full distribution of the same information, using transparency for the overlaps
-   pH distributions were observed using both faceted density plots and boxplots. These visualizations showed the significant changes in pH over the seasons in 2020 and 2021. These are interestingly very different across both years.

## Variable Time Series and Daily Averages

-   Daily averages were calculated using `group_by()` and `summarize()`. Orignal measurements were collected every 15 minutes. This was used to reduce noise/ fluctuations and highlight broader trends.
-   Line plots showed seasonal patterns for pH, salinity, temperature, depth, conductivity, turbidity, chlorophyll fluorescence, and dissolved oxygen
-   `Geom_smooth()` was used to add trend lines

## Correlation Plots (Using Means)

-   Scatterplots used `geom_jitter` to make plots more easily interpretable
-   Facets by season showed differences in the strength of correlations
-   Salinity was mutated into bins to look at mean temperature against mean salinity
-   Salinity plots were replicated using the Conductivity variable instead in order to return to a raw measurement (salinity is in part derived from conductivity)
-   Plots were also used to verify relationship between salinity and conductivity

## Correlation Plots

```{r, result = FALSE, eval=FALSE}
daily_2020 <- no_winter_2020 %>%
  group_by(date_clean, season)%>%
  summarise(
    meanpH = mean(pH, na.rm = TRUE),
    meanSal = mean(Sal, na.rm = TRUE),
    meanTemp = mean(Temp, na.rm = TRUE),
    meanDepth = mean(Depth, na.rm = TRUE),
    meanSpCond = mean(SpCond, na.rm = TRUE),
    meanTurb = mean(Turb, na.rm = TRUE),
    meanChlFluor = mean(ChlFluor, na.rm = TRUE),
    meanDO_Pct = mean(DO_Pct, na.rm = TRUE),
    meanDO_mgl = mean(DO_mgl, na.rm = TRUE))

daily_2021 <- no_winter_2021 %>%
  group_by(date_clean, season)%>%
  summarise(
    meanpH = mean(pH, na.rm = TRUE),
    meanSal = mean(Sal, na.rm = TRUE),
    meanTemp = mean(Temp, na.rm = TRUE),
    meanDepth = mean(Depth, na.rm = TRUE),
    meanSpCond = mean(SpCond, na.rm = TRUE),
    meanTurb = mean(Turb, na.rm = TRUE),
    meanChlFluor = mean(ChlFluor, na.rm = TRUE),
    meanDO_Pct = mean(DO_Pct, na.rm = TRUE),
    meanDO_mgl = mean(DO_mgl, na.rm = TRUE))


```

```{r, result = FALSE, eval=FALSE}
ggplot(daily_2020, aes(x = meanpH, y = meanSal))+
  geom_jitter(alpha = 0.5, color = "royalblue3") +
  geom_smooth(method = "lm", se = FALSE, color = "black")+
  scale_x_continuous(limits = c(6.5, 8.5)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Mean pH v. Mean Salinity in 2020",
       x = "Mean pH",
       y = "Mean Salinity")
```

-   Not very nice plot!

## Mean pH v. Mean Salinity in 2021

```{r, result = FALSE, eval=FALSE}
ggplot(daily_2021, aes(x = meanpH, y = meanSal))+
  geom_jitter(alpha = 0.5, color = "mediumorchid3") +
  geom_smooth(method = "lm", se = FALSE, color = "black")+ 
  scale_x_continuous(limits = c(6.5, 8.5)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Mean pH v. Mean Salinity in 2021",
       x = "Mean pH",
       y = "Mean Salinity")
```

-   Also not very nice

## Mean pH v. Mean Conductivity 2020

```{r, result = FALSE, eval=FALSE}
ggplot(daily_2020, aes(x = meanpH, y = meanSpCond))+
  geom_jitter(alpha = 0.5, color = "royalblue3") +
  geom_smooth(method = "lm", se = FALSE, color = "black")+ 
  scale_x_continuous(limits = c(6.5, 8.5)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Mean pH v. Mean Conductivity in 2020",
       x = "Mean pH",
       y = "Mean Conductivity")
```

-   Much nicer

## Mean pH v. Mean Conductivity 2021

```{r, result = FALSE, eval=FALSE}
ggplot(daily_2021, aes(x = meanpH, y = meanSpCond))+
  geom_jitter(alpha = 0.5, color = "mediumorchid3") +
  geom_smooth(method = "lm", se = FALSE, color = "black")+ 
  scale_x_continuous(limits = c(6.5, 8.5)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Mean pH v. Mean Conductivity in 2021",
       x = "Mean pH",
       y = "Mean Conductivity")
```

-   Also nicer!

## Combined Dataset Exploration

-   

    ```         
    Later, I combined both datasets in order to closely compared both years for former plots
    ```

```{r, result = FALSE, eval=FALSE}
no_winter_year_2020 <- no_winter_2020 %>% mutate(year = 2020)
no_winter_year_2021 <- no_winter_2021 %>% mutate(year = 2021)
combined_datasets <- bind_rows(no_winter_year_2020, no_winter_year_2021)

combined_datasets %>%
  pivot_longer(
    cols = c(Temp, Sal, Turb, ChlFluor, Level, Depth, pH, SpCond, DO_Pct, DO_mgl),
    names_to = "variable",
    values_to = "value") %>%
  ggplot(aes(x = factor(year), y = value, fill = factor(year)))+
  geom_boxplot()+
  facet_wrap(~variable, scales = "free_y")+
  labs(y = NULL)+
  labs(title = "Distribution of Variables in 2020 and 2021",
       x = "Year",
       y = "Value")+
  scale_fill_manual(values = c("2020" = "royalblue3", "2021" = "mediumorchid3"))
```

## And This:

```{r, result = FALSE, eval=FALSE}

daily_combined_datasets <- combined_datasets %>%
  group_by(year, month, date_clean, season)%>%
  summarise(
    meanpH = mean(pH, na.rm = TRUE),
    meanSal = mean(Sal, na.rm = TRUE),
    meanTemp = mean(Temp, na.rm = TRUE),
    meanDepth = mean(Depth, na.rm = TRUE),
    meanSpCond = mean(SpCond, na.rm = TRUE),
    meanTurb = mean(Turb, na.rm = TRUE),
    meanChlFluor = mean(ChlFluor, na.rm = TRUE),
    meanDO_Pct = mean(DO_Pct, na.rm = TRUE),
    meanDO_mgl = mean(DO_mgl, na.rm = TRUE),
    .groups = "drop"
)

daily_combined_datasets$doy <-
  as.numeric(format(daily_combined_datasets$date_clean, "%j"))

ggplot(daily_combined_datasets, aes(x = meanTemp, y = meanpH, color = factor(year), group = factor(year)))+
  geom_jitter(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE)+
  labs(title = "Mean Temperature v. Mean pH",
       x = "Mean Temperature",
       y = "Mean pH")+
  scale_color_manual(values = c("2020" = "royalblue3", "2021" = "mediumorchid3"))
```

## Also this:

```{r fig.width=16, fig.height=4, eval = FALSE}
#figure measurements edited in order to better fit plot
ggplot(combined_datasets, aes(x = as.integer(format(date_clean, "%d")), y = Temp, color = factor(year), group = factor(year)))+
  geom_smooth(method = "loess", se = FALSE)+
  facet_wrap(~ month, scales = "free_x", nrow = 1) +
  labs(title = "Mean Temperature by Month in 2020 and 2021",
       x = "Date",
       y = "Temp")+
  scale_color_manual(values = c("2020" = "royalblue3", "2021" = "mediumorchid3"))

```

-   

    ```         
     Plots were made from daily means for each variable over the course of the year for 2020 and 2021
    ```

-   

    ```         
     Smoothed lines for better interpretability of broad trend.
    ```

-   

    ```         
     The combined dataset works more effectively for comparison
    ```

-   

    ```         
      Also better for consistency for x and y axis
    ```

## Conclusion

-   

    ```         
     Analysis of the data revealed interesting distributional differences between water quality indicators/ variables
    ```

-   

    ```         
     Differences between season, month, and year were also discovered
    ```

-   

    ```         
     Missing data in 2021 may limit findings
    ```

-   In Future:

-   

    ```         
     Add more years for stronger trends
    ```

-   

    ```         
     Link water quality shifts to external conditions such as climate or tides
    ```

-   

    ```         
     Create model for prediction of these variables
    ```
