---
title: "stony_creek_tables"
author: Natalie Novoseltseva
format: html
---
Measuring Climate Change Effects through Water Quality 
Stony Creek EDA

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#libraries
library(tidyverse)
library(dplyr)
library(lubridate)
library(gt)
library(ggplot2)
library(stringr)
library(janitor)
```

## Data


```{r}
#stony creek data 2002-2021

#creates a data folder for the specific site
dir.create("data/stony_creek", recursive = TRUE, showWarnings = FALSE) 

#using tempfile() to create a tempary storage
zip <- tempfile(fileext = ".zip")

#downlod the file from the url
download.file("http://cdmo.baruch.sc.edu/aqs/output/60521.zip", zip, mode = "wb")

unzip(zip, exdir = "data/stony_creek")

#read in the files for all the years as a one data file
files <- list.files(
  "data/stony_creek",
  pattern = "^hudscwq[0-9]{4}\\.csv$", # ^ -starts a string
                                       # hudscwq - the exact name in the file (constant)
                                       # any digits from 0-9 might follow
                                       # {4} - we are looking for 4 digits pattern
                                       # \\. - there should be a dot
                                       # csv - it should end with "csv"
                                       # end a string
  full.names = TRUE
)

#combines all data files into one dataset
sc_02_21_raw <- do.call(
  rbind,
  lapply(files, function(f) {
    df <- read.csv(f, fileEncoding = "Latin1")
    df$year <- as.integer(gsub(".*([0-9]{4})\\.csv$", "\\1", f))
    df
  })
)
```

## Data Cleaning

```{r}
# Removing empty columns and other irrelevant columns
stony_creek <- sc_02_21_raw |>
  remove_empty("cols")|>
  select(-StationCode,
        -isSWMP, #historical and provisionalPlus columns indicate the status of QAQC
        -Historical, 
        -ProvisionalPlus
        
  )

#creating a vector with all the flags
flags <- c("F_Record", "F_Temp", "F_SpCond", "F_Sal", "F_DO_Pct", "F_Depth", "F_cDepth", "F_Level", "F_pH", "F_Turb", "F_ChlFluor", "F_DO_mgl")

stony_creek <- stony_creek |>
  mutate(Temp = if_else(
      if_any(all_of(flags), ~ str_detect(.x, "<-3>|<-2>|<1>")), NA_real_, Temp))

#removing flag columns #remove comment of the next two lines once the previous step was done
# stony_creek <- stony_creek |>
#   select(-flags) 

```


```{r}
#separate the data and time column
sc_02_21 <- sc_02_21_raw %>%
  mutate(
    DateTimeStamp = as.POSIXct(DateTimeStamp, format = "%m/%d/%Y %H:%M"), #converts the variable into standard POSIXct format
    date = as.Date(DateTimeStamp),
    time = format(DateTimeStamp, "%H:%M:%S")
  )

#create a month column
sc_02_21 <- sc_02_21 %>%
  mutate(
    month = month(DateTimeStamp)          
  )

#create a season column
sc_02_21 <- sc_02_21 %>%
  mutate(
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% 3:5        ~ "Spring",
      month %in% 6:8        ~ "Summer",
      month %in% 9:11       ~ "Fall"
    )
  )
```


## Tables and averages

Number of observations

```{r}
#extract observations with missing season values
sc_02_21_NAs <- sc_02_21 %>%
  filter(is.na(season))

#count observations by year and season and display as a formatted table
sc_02_21 %>%
  count(year, season) %>%
  gt(groupname_col = "season") %>%
  tab_header(
    title = "Observation Coverage by Year and Season"
  ) %>%
  cols_label(
    year = "Year",
    n = "Number of Observations"
  )

```

Summary statistics (all years)

```{r}
#compute summary statistics for selected variables, reshape, and display as a formatted table
sc_02_21 %>%
  summarise(
    Temp_mean = mean(Temp, na.rm = TRUE),
    Temp_sd   = sd(Temp, na.rm = TRUE),
    Temp_min  = min(Temp, na.rm = TRUE),
    Temp_max  = max(Temp, na.rm = TRUE),

    DO_mean   = mean(DO_mgl, na.rm = TRUE),
    pH_mean   = mean(pH, na.rm = TRUE),
    Sal_mean  = mean(Sal, na.rm = TRUE)
  ) %>%
#convert wide summary to long format for easier display
  pivot_longer(everything(),
               names_to = "Statistic",
               values_to = "Value") %>%
#format as a table using gt
  gt() %>%
  tab_header(
    title = "Overall Summary Statistics"
  )
```

Averages by season

```{r}
sc_02_21 %>%
  group_by(season) %>%
  summarise(
    avg_temp = mean(Temp, na.rm = TRUE),
    avg_do   = mean(DO_mgl, na.rm = TRUE),
    avg_ph   = mean(pH, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  gt() %>%
  tab_header(
    title = "Average Water Quality Measures by Season"
  ) %>%
  fmt_number(decimals = 2)
```

Average temperatures by year

```{r}
#calculate average water quality measures by season and display as a formatted table
avg_temp_by_year <- sc_02_21 %>%
  group_by(year) %>%
  summarise(
    avg_temp = mean(Temp, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )
#format the summary as a table using gt
avg_temp_by_year %>%
  gt() %>%
  tab_header(
    title = "Average Water Temperature by Year"
  ) %>%
  fmt_number(
    columns = avg_temp,
    decimals = 2
  ) %>%
  cols_label(
    year = "Year",
    avg_temp = "Average Temperature (°C)",
    n_obs = "Observations"
  )
```

Average temperature by year and season (one table)

```{r}
#calculate average temperature and number of observations by season and year
avg_temp_by_year_season <- sc_02_21 %>%
  group_by(season, year) %>%
  summarise(
    avg_temp = mean(Temp, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  ) %>%
#remove rows with missing season values
  filter(!is.na(season))

#display the results as a formatted table
avg_temp_by_year_season %>%
  gt(groupname_col = "season") %>%
  tab_header(
    title = "Average Water Temperature by Year and Season"
  ) %>%
#round average temperature to 2 decimal places
  fmt_number(
    columns = avg_temp,
    decimals = 2
  ) %>%
#rename columns
  cols_label(
    year = "Year",
    avg_temp = "Average Temperature (°C)",
    n_obs = "Observations"
  )
```

Average temperature by year and season (as separate tables)

```{r}
#split the data by season and create a separate formatted table for each season
season_tables <- avg_temp_by_year_season %>%
  split(.$season) %>%
  lapply(function(season_df) {
    #get the season name
    season_name <- unique(season_df$season)

#select relevant columns and format as a gt table
    season_df %>%
      select(year, avg_temp, n_obs) %>%
      gt() %>%
      tab_header(
        title = paste("Average Water Temperature by Year"),
        subtitle = paste("Season:", season_name)
      ) %>%
      fmt_number(
        columns = avg_temp,
        decimals = 2
      ) %>%
      cols_label(
        year = "Year",
        avg_temp = "Average Temperature (°C)",
        n_obs = "Observations"
      )
  })

#access the tables for each season
season_tables$Spring
season_tables$Summer
season_tables$Fall
season_tables$Winter
```

Seasonal extremes

```{r}
#calculate seasonal temperature extremes and display as a formatted table
sc_02_21 %>%
  group_by(season) %>%
  summarise(
    min_temp = min(Temp, na.rm = TRUE), #lowest observed temperature
    max_temp = max(Temp, na.rm = TRUE), #highest observed temperature
    p95_temp = quantile(Temp, 0.95, na.rm = TRUE), #the 95th percentile of temperature for each season, which excludes outliers (5% of the highest observation)
    .groups = "drop"
  ) %>%
#format the summary as a table
  gt() %>%
  tab_header(
    title = "Seasonal Temperature Extremes"
  )
```

Changes in the magnitude of temperature changes

```{r}
#calculate yearly average air temperature and its deviation from the overall mean
avg_air_temp_by_year <- sc_02_21 %>%
  group_by(year) %>%
  summarise(
    avg_air_temp = mean(Temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    delta_overall = avg_air_temp - mean(avg_air_temp, na.rm = TRUE)  #difference from overall mean
  )

#gt table
avg_air_temp_by_year %>%
  gt() %>%
  tab_header(
    title = "Average Air Temperature by Year",
    subtitle = "Deviation from Overall Mean (2001–2026)"
  ) %>%
  fmt_number(
    columns = c(avg_air_temp, delta_overall),
    decimals = 2
  ) %>%
  cols_label(
    year = "Year",
    avg_air_temp = "Average Air Temperature (°C)",
    delta_overall = "Difference from Overall Average (°C)"
  )
```

Plot of the of the magnitude of temperature changes

```{r}
#calculate yearly average air temperature and deviation from overall mean
avg_air_temp_by_year <- sc_02_21 %>%
  group_by(year) %>%
  summarise(
    avg_air_temp = mean(Temp, na.rm = TRUE), #yearly average temperature
    .groups = "drop"
  ) %>%
  mutate(
    delta_overall = avg_air_temp - mean(avg_air_temp, na.rm = TRUE)  #deviation from overall mean
  )

#store overall mean temperature for reference in the plot
overall_mean <- mean(avg_air_temp_by_year$avg_air_temp)

#plot
ggplot(avg_air_temp_by_year, aes(x = factor(year), y = avg_air_temp, fill = delta_overall > 0)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = overall_mean, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Average Air Temperature by Year",
    subtitle = "Bars show yearly averages; red dashed line = overall mean (2001–2026)",
    x = "Year",
    y = "Average Air Temperature (°C)"
  ) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "lightgray")) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Changes in the magnitude of temperature changes

```{r}
#calculate yearly average temperature and year-to-year changes
sc_02_21 %>%
sc_02_21 %>%
  group_by(year) %>%
  summarise(
    avg_temp = mean(Temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    delta_temp = avg_temp - lag(avg_temp)
  ) %>%
#display as a table
  gt() %>%
  tab_header(
    title = "Year-to-Year Change in Average Temperature"
  )
```

Plot of the of the magnitude of temperature changes

```{r}
#calculate yearly average temperature and year-to-year changes
yearly_temp_change <- sc_02_21 %>%
  group_by(year) %>%
  summarise(
    avg_temp = mean(Temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year) %>% #chronological order
  mutate(
    delta_temp = avg_temp - lag(avg_temp) #change from the previous year
  )
#plot year-to-year temperature changes as a bar chart
ggplot(yearly_temp_change, aes(x = year, y = delta_temp, fill = delta_temp > 0)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Year-to-Year Change in Average Temperature",
    x = "Year",
    y = "Change in Average Temperature (°C)"
  ) +
  theme_minimal()
```

Joining the water quality and weather datasets (?)

```{r}
#merged_data <- left_join(
#  sc_02_21,
#  fs_weather_01_26,
#  by = c("DateTimeStamp" = "DatetimeStamp")
#)
```

