---
author: Ilianna Ditren
---

# Climate Change Effects of Water Quality (20-years change)

## Project Set Up

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#Downloading libraries

library(readr)
library(tidyverse)
library(dplyr)
library(lubridate)
library(janitor)

```

## Downloading and Data Cleaning

```{r, echo = FALSE}

#Stony Creek Data 2002-2021
dir.create("data/stony_creek", recursive = TRUE, showWarnings = FALSE) 
#creates a folder for the new data

#downloading Stony Creek data file from the web
zip_sc <- tempfile(fileext = ".zip")                                       #creating a temporary zip file

stony_creek_raw <- "http://cdmo.baruch.sc.edu/aqs/output/60521.zip"

download.file(stony_creek_raw, zip_sc, mode = "wb")

unzip(zip_sc, exdir = "data/stony_creek")                            
#unzipping data from and placing it in a folder
```

```{r, echo = FALSE}
#read in the files for all the years as a one data file
files <- list.files(
  "data/stony_creek",
  pattern = "^hudscwq[0-9]{4}\\.csv$", # ^ -starts a string, hudscwq - the exact name in the file (constant), any digits from 0-9 might follow, {4} - we are looking for 4 digits pattern, \\. - there should be a dot, csv - it should end with "csv",  end a string
  full.names = TRUE
)

sc_02_21_raw <- do.call(
  rbind,
  lapply(files, function(f) {
    df <- read.csv(f, fileEncoding = "Latin1")
    df$year <- as.integer(gsub(".*([0-9]{4})\\.csv$", "\\1", f))
    df
  })
)

```

```{r, eval = FALSE}

# Removing empty columns and other irrelant columns
t <- sc_02_21_raw |>
  remove_empty("cols")|>
  select(-StationCode,
        -isSWMP, #historical and provisionalPlus columns indicate the status of QAQC
        -Historical, 
        -ProvisionalPlus,
  )

#creating a vector with all the flags
flags <- c("F_Record", "F_Temp", "F_SpCond", "F_Sal", "F_DO_Pct", "F_Depth", "F_cDepth", "F_Level", "F_pH", "F_Turb", "F_ChlFluor", "F_DO_mgl")

t <- t |>
  mutate(Temp = if_else(
      if_any(all_of(flags), ~ str_detect(.x, "<-3>|<-2>|<1>")), NA_real_, Temp))

#removing flag columns #remove comment of the next two lines once the previous step was done
# stony_creek <- stony_creek |>
#   select(-flags) 


```

```{r}

# Removing empty columns and other irrelant columns
stony_creek <- sc_02_21_raw |>
  remove_empty("cols")|>
  select(-StationCode,
        -isSWMP, #historical and provisional Plus columns indicate the status of QAQC
        -Historical, 
        -ProvisionalPlus,
  )

#creating a vector with all the flags
flags <- c("F_Record", "F_Temp", "F_SpCond", "F_Sal", "F_DO_Pct", "F_Depth", "F_cDepth", "F_Level", "F_pH", "F_Turb", "F_ChlFluor", "F_DO_mgl")

stony_creek <- stony_creek |>
  mutate(Temp = if_else(str_detect(F_Temp, "<-3>|<-2>|<1>"), NA_real_, Temp),
         SpCond = if_else(str_detect(F_SpCond, "<-3>|<-2>|<1>"), NA_real_, SpCond),
         Sal = if_else(str_detect(F_Sal, "<-3>|<-2>|<1>"), NA_real_, Sal),
         DO_Pct = if_else(str_detect(F_DO_Pct, "<-3>|<-2>|<1>"), NA_real_, DO_Pct),
         Depth = if_else(str_detect(F_Depth, "<-3>|<-2>|<1>"), NA_real_, Depth),
         cDepth = if_else(str_detect(F_cDepth, "<-3>|<-2>|<1>"), NA_real_, cDepth),
         pH = if_else(str_detect(F_pH, "<-3>|<-2>|<1>"), NA_real_, pH),
         Turb = if_else(str_detect(F_Turb, "<-3>|<-2>|<1>"), NA_real_, Turb),
         ChlFluor = if_else(str_detect(F_ChlFluor, "<-3>|<-2>|<1>"), NA_real_, ChlFluor),
         DO_mgl = if_else(str_detect(DO_mgl, "<-3>|<-2>|<1>"), NA_real_, DO_mgl)
  )

#removing flag columns 
stony_creek <- stony_creek %>% select(-all_of(flags))

```


```{r, echo = FALSE}
#separate the date and time column
stony_creek <- stony_creek %>%
  mutate(
    DateTimeStamp = as.POSIXct(DateTimeStamp, format = "%m/%d/%Y %H:%M"), #converts the variable into standard POSIXct format
    date = as.Date(DateTimeStamp),
    time = format(DateTimeStamp, "%H:%M:%S")
  )

#create a month column
stony_creek <- stony_creek %>%
  mutate(
    month = month(DateTimeStamp)          
  )

#create a season column
stony_creek <- stony_creek %>%
  mutate(
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% 3:5        ~ "Spring",
      month %in% 6:8        ~ "Summer",
      month %in% 9:11       ~ "Fall"
    )
  )

```

## Exploratory Data Analysis

### Water Temperature - Temp

This section will begin by looking at water temperatures by season across the years 2002 - 2022.

```{r, echo = FALSE, message = FALSE}

#Selecting relevant data to look at temperature
temp <- stony_creek |>
  select(Temp, year, season)|>
  filter(!is.na(Temp))

#Creating mean column
mean_temp <- temp |> 
  group_by(season, year)|>
  summarize(mean_temp = mean(Temp))

```

```{r, echo =FALSE}
#Creating plots by the season
##Spring temp
mean_spring <- mean_temp |>
  filter(season == "Spring")

line_mean_spring <- ggplot(mean_spring)+
  geom_point(aes(year, mean_temp))+
  labs(title = "Mean Spring Temperature by Year")

line_mean_spring

```

```{r, echo=FALSE, message=FALSE}
##Fall temp
mean_fall <- mean_temp |>
  filter(season == "Fall")

line_mean_fall <- ggplot(mean_fall)+
  geom_point(aes(year, mean_temp))+
  labs(title = "Mean Fall Temperature by Year")

line_mean_fall

```

```{r, echo=FALSE, message=FALSE}
##Summer temp
mean_summer <- mean_temp |>
  filter(season == "Summer")

line_mean_summer <- ggplot(mean_summer)+
  geom_point(aes(year, mean_temp))+
  labs(title = "Mean Summer Temperature by Year")

line_mean_summer

```

```{r, echo=FALSE, message=FALSE}
##Winter
mean_winter <- mean_temp |>
  filter(season == "Winter")

line_mean_winter <- ggplot(mean_winter)+
  geom_point(aes(year, mean_temp))+
  labs(title = "Mean Winter Temperature by Year")

line_mean_winter
```

```{r, echo=FALSE, message=FALSE}
#plotting temperature by all seasons

seasons_temp <- mean_temp|> filter(!is.na(season))|> 
  #removed 6 rows with missing values
  ggplot()+
  geom_point(aes(year, mean_temp, color = season))+
  scale_color_viridis_d()+
  labs(title = "Mean Seasonal Temperature by Year")

seasons_temp

```

# potential of Hydrogen - (pH)

This section will begin by looking at the water ph by season across the years 2002 - 2022.

```{r, echo = FALSE, message = FALSE}

#Selecting relevant data to look at water pH
ph <- stony_creek |>
  select(pH, year, season)|>
  filter(!is.na(pH))

#Creating mean column
mean_ph <- ph |> 
  group_by(season, year)|>
  summarize(mean_ph = mean(pH))

```

```{r, echo=FALSE, message=FALSE}
#Creating plots by the season
##Spring ph
ph_spring <- mean_ph |>
  filter(season == "Spring")

line_ph_spring <- ggplot(ph_spring)+
  geom_point(aes(year, mean_ph))+
  labs(title = "Mean Spring pH Level by Year")

line_ph_spring
```

```{r, echo=FALSE, message=FALSE}
##Fall ph
ph_fall <- mean_ph |>
  filter(season == "Fall")

line_ph_fall <- ggplot(ph_fall)+
  geom_point(aes(year, mean_ph))+
  labs(title = "Mean Fall pH Level by Year")

line_ph_fall
```

```{r, echo=FALSE, message=FALSE}
##Summer ph
ph_summer <- mean_ph |>
  filter(season == "Summer")

line_ph_summer <- ggplot(ph_summer)+
  geom_point(aes(year, mean_ph))+
  labs(title = "Mean Summer pH Level by Year")

line_ph_summer
```

```{r, echo=FALSE, message=FALSE}
##Winter pH Levels
ph_winter <- mean_ph |>
  filter(season == "Winter")

line_ph_winter <- ggplot(ph_winter)+
  geom_point(aes(year, mean_ph))+
  labs(title = "Mean Winter pH Level by Year")

line_ph_winter
```

```{r, echo=FALSE, message=FALSE}
#plotting pH Levels by all seasons

seasons_ph <- mean_ph|> filter(!is.na(season))|>
  ggplot()+
  geom_point(aes(year, mean_ph, color = season))+
  scale_color_viridis_d()+
  labs(title = "Mean Seasonal pH Levels by Year")

seasons_ph

```

#Turbity of Water - (Turb)

This section will begin by looking at the water turbidity by season across the years 2002 - 2022.

```{r, echo = FALSE, message = FALSE}

#Selecting relevant data to look at water turbidity only
turb <- stony_creek |>
  select(Turb, year, season)|>
  filter(!is.na(Turb))

#Creating mean column
mean_turb <- turb |> 
  group_by(season, year)|>
  summarize(mean_turb = mean(Turb))

```

```{r, echo=FALSE, message=FALSE}
#Creating plots by the season
##Spring turbidity
turb_spring <- mean_turb |>
  filter(season == "Spring")

line_turb_spring <- ggplot(turb_spring)+
  geom_point(aes(year, mean_turb))+
  labs(title = "Mean Spring Water Turbidity Level by Year")

line_turb_spring
```
Turbidity levels in Spring varies significantly every year. I wonder if that has anything to do with varying precipitation those years.

```{r, echo=FALSE, message=FALSE}
##Fall turbidity
turb_fall <- mean_turb |>
  filter(season == "Fall")

line_turb_fall <- ggplot(turb_fall)+
  geom_point(aes(year, mean_turb))+
  labs(title = "Mean Fall Water Turbidity Level by Year")

line_turb_fall
```

```{r, echo=FALSE, message=FALSE}
##Summer turbidity
turb_summer <- mean_turb |>
  filter(season == "Summer")

line_turb_summer <- ggplot(turb_summer)+
  geom_point(aes(year, mean_turb))+
  labs(title = "Mean Summer Water Turbity Level by Year")

line_turb_summer
```

```{r, echo=FALSE, message=FALSE}
##Winter Water Turbidity
turb_winter <- mean_turb |>
  filter(season == "Winter")

line_turb_winter <- ggplot(turb_winter)+
  geom_point(aes(year, mean_turb))+
  labs(title = "Mean Winter Water Turbidity Level by Year")

line_turb_winter
```

```{r, echo=FALSE, message=FALSE}
#plotting Water Turbidity by all seasons

seasons_turb <- mean_turb|> filter(!is.na(season))|>
  ggplot()+
  geom_point(aes(year, mean_turb, color = season))+
  scale_color_viridis_d()+
  labs(title = "Mean Seasonal Water Turbidity Levels by Year")

seasons_turb

```

#Dissolved oxygen measured in percent saturation (%) - (DO_pct)

This section will begin by looking at the dissolved oxygen by season across the years 2002 - 2022.

```{r, echo = FALSE, message = FALSE}

#Selecting relevant data to look at Dissolved Oxygen Percent only
do_pct <- stony_creek |>
  select(DO_Pct, year, season)|>
  filter(!is.na(DO_Pct))

#Creating mean column
mean_do_pct <- do_pct |> 
  group_by(season, year)|>
  summarize(mean_do_pct = mean(DO_Pct))

```

```{r, echo=FALSE, message=FALSE}
#Creating plots by the season
##Spring Dissolved Oxygen Percent
do_pct_spring <- mean_do_pct |>
  filter(season == "Spring")

line_dopct_spring <- ggplot(do_pct_spring)+
  geom_point(aes(year, mean_do_pct))+
  labs(title = "Mean Spring Dissolved Oxygent Percent Saturation Level by Year")

line_dopct_spring
```

```{r, echo=FALSE, message=FALSE}
##Fall Dissolved Oxygen Percent
do_pct_fall <- mean_do_pct |>
  filter(season == "Fall")

line_dopct_fall <- ggplot(do_pct_fall)+
  geom_point(aes(year, mean_do_pct))+
  labs(title = "Mean Fall Dissolved Oxygen Percent Saturation Level by Year")

line_dopct_fall
```

```{r, echo=FALSE, message=FALSE}
##Summer Dissolved Oxygen Percent
do_pct_summer <- mean_do_pct |>
  filter(season == "Summer")

line_dopct_summer <- ggplot(do_pct_summer)+
  geom_point(aes(year, mean_do_pct))+
  labs(title = "Mean Summer Dissolved Oxygen Percent Saturation Level by Year")

line_dopct_summer
```

```{r, echo=FALSE, message=FALSE}
##Winter Dissolved Oxygen Percent
do_pct_winter <- mean_do_pct |>
  filter(season == "Winter")

line_dopct_winter <- ggplot(do_pct_winter)+
  geom_point(aes(year, mean_do_pct))+
  labs(title = "Mean Winter Dissolved Oxygen Percent Saturation Level by Year")

line_dopct_winter
```

```{r, echo=FALSE, message=FALSE}
#plotting temperature by all seasons

seasons_do_pct <- mean_do_pct|> filter(!is.na(season))|>
  ggplot()+
  geom_point(aes(year, mean_do_pct, color = season))+
  scale_color_viridis_d()+
  labs(title = "Mean Seasonal Dissolved Oxygen Percent Saturation Level by Year")

seasons_do_pct

```

#Dissolved oxygen concentration measured in milligrams per Liter (mg/L) - (DO_mgl)

This section will begin by looking at the dissolved oxygen by season across the years 2002 - 2022.

```{r, echo = FALSE, message = FALSE}

#Selecting relevant data to look at Dissolved Oxygen concentration only
do_mgl <- stony_creek |>
  select(DO_mgl, year, season)|>
  filter(!is.na(DO_mgl))

#Creating mean column
mean_do_mgl <- do_mgl |> 
  group_by(season, year)|>
  summarize(mean_do_mgl = mean(DO_mgl))

```

```{r, echo=FALSE, message=FALSE}
#Creating plots by the season
##Spring Dissolved Oxygen concentration
do_mgl_spring <- mean_do_mgl |>
  filter(season == "Spring")

line_domgl_spring <- ggplot(do_mgl_spring)+
  geom_point(aes(year, mean_do_mgl))+
  labs(title = "Mean Spring Dissolved Oxygen concentration 
       measured in milligrams per Liter (mg/L) by Year")

line_domgl_spring
```

```{r, echo=FALSE, message=FALSE}
##Fall Dissolved Oxygen concentration
do_mgl_fall <- mean_do_mgl |>
  filter(season == "Fall")

line_domgl_fall <- ggplot(do_mgl_fall)+
  geom_point(aes(year, mean_do_mgl))+
  labs(title = "Mean Fall Dissolved Oxygen concentration 
       measured in milligrams per Liter (mg/L) by Year")

line_domgl_fall
```

```{r, echo=FALSE, message=FALSE}
##Summer Dissolved Oxygen concentration 
do_mgl_summer <- mean_do_mgl |>
  filter(season == "Summer")

line_domgl_summer <- ggplot(do_mgl_summer)+
  geom_point(aes(year, mean_do_mgl))+
  labs(title = "Mean Summer Dissolved Oxygen concentration 
       measured in milligrams per Liter (mg/L) by Year")

line_domgl_summer
```

```{r, echo=FALSE, message=FALSE}
##Winter Dissolved Oxygen concentration
do_pct_winter <- mean_do_pct |>
  filter(season == "Winter")

line_dopct_winter <- ggplot(do_pct_winter)+
  geom_point(aes(year, mean_do_pct))+
  labs(title = "Mean Winter Dissolved Oxygen concentration 
       measured in milligrams per Liter (mg/L) by Year")

line_dopct_winter
```

```{r, echo=FALSE, message=FALSE}
#plotting Dissolved Oxygen concentration by all seasons

seasons_do_mgl <- mean_do_mgl|> filter(!is.na(season))|>
  ggplot()+
  geom_point(aes(year, mean_do_mgl, color = season))+
  scale_color_viridis_d()+
  labs(title = "Mean Seasonal Dissolved Oxygen concentration 
       measured in milligrams per Liter (mg/L) by Year")

seasons_do_mgl

```


#Water Conductivity (sp_cond)

This section will begin by looking at water conductivity by season across the years 2002 - 2022.

```{r, echo = FALSE, message = FALSE}

#Selecting relevant data to look at water conductivity only
sp_cond <- stony_creek |>
  select(SpCond, year, season)|>
  filter(!is.na(SpCond))

#Creating mean column
mean_sp_cond <- sp_cond |> 
  group_by(season, year)|>
  summarize(mean_sp_cond = mean(SpCond))

```

```{r, echo=FALSE, message=FALSE}
#Creating plots by the season
##Spring water conductivity
sp_cond_spring <- mean_sp_cond |>
  filter(season == "Spring")

line_spcond_spring <- ggplot(sp_cond_spring)+
  geom_point(aes(year, mean_sp_cond))+
  labs(title = "Mean Spring Water Conductivity by Year")

line_spcond_spring
```

```{r, echo=FALSE, message=FALSE}
##Fall water conductivity
sp_cond_fall <- mean_sp_cond |>
  filter(season == "Fall")

line_spcond_fall <- ggplot(sp_cond_fall)+
  geom_point(aes(year, mean_sp_cond))+
  labs(title = "Mean Fall Water Conductivity by Year")

line_spcond_fall
```

```{r, echo=FALSE, message=FALSE}
##Summer water conductivity
sp_cond_summer <- mean_sp_cond |>
  filter(season == "Summer")

line_spcond_summer <- ggplot(sp_cond_summer)+
  geom_point(aes(year, mean_sp_cond))+
  labs(title = "Mean Summer Water Conductivity by Year")

line_spcond_summer
```

```{r, echo=FALSE, message=FALSE}
##Water conductivity for Winter
sp_cond_winter <- mean_sp_cond |>
  filter(season == "Winter")

line_spcond_winter <- ggplot(sp_cond_winter)+
  geom_point(aes(year, mean_sp_cond))+
  labs(title = "Mean Winter Water Conductivity by Year")

line_spcond_winter
```

```{r, echo=FALSE, message=FALSE}
#plotting water conductivity by all seasons

seasons_sp_cond <- mean_sp_cond|> filter(!is.na(season))|>
  ggplot()+
  geom_point(aes(year, mean_sp_cond, color = season))+
  scale_color_viridis_d()+
  labs(title = "Mean Seasonal Water Conductivity by Year")

seasons_sp_cond

```

#Salinity (sal)

This section will begin by looking at the water salinity by season across the years 2002 - 2022.

```{r, echo = FALSE, message = FALSE}

#Selecting relevant data to look at water salinity only
sal <- stony_creek |>
  select(Sal, year, season)|>
  filter(!is.na(Sal))

#Creating mean column
mean_sal <- sal |> 
  group_by(season, year)|>
  summarize(mean_sal = mean(Sal))

```

```{r, echo=FALSE, message=FALSE}
#Creating plots by the season
##Spring salinity
sal_spring <- mean_sal |>
  filter(season == "Spring")

line_sal_spring <- ggplot(sal_spring)+
  geom_point(aes(year, mean_sal))+
  labs(title = "Mean Spring Water Salinity by Year")

line_sal_spring
```

```{r, echo=FALSE, message=FALSE}
##Fall salinity
sal_fall <- mean_sal |>
  filter(season == "Fall")

line_sal_fall <- ggplot(sal_fall)+
  geom_point(aes(year, mean_sal))+
  labs(title = "Mean Fall Water Salinity by Year")

line_sal_fall
```

```{r, echo=FALSE, message=FALSE}
##Summer water salinity
sal_summer <- mean_sal |>
  filter(season == "Summer")

line_sal_summer <- ggplot(sal_summer)+
  geom_point(aes(year, mean_sal))+
  labs(title = "Mean Summer Water Salinity by Year")

line_sal_summer
```

```{r, echo=FALSE, message=FALSE}
##Winter Water Salinity
sal_winter <- mean_sal |>
  filter(season == "Winter")

line_sal_winter <- ggplot(sal_winter)+
  geom_point(aes(year, mean_sal))+
  labs(title = "Mean Winter Water Salinity by Year")

line_sal_winter
```

```{r, echo=FALSE, message=FALSE}
#plotting Water Salinity by all seasons

seasons_sal <- mean_sal|> filter(!is.na(season))|>
  ggplot()+
  geom_point(aes(year, mean_sal, color = season))+
  scale_color_viridis_d()+
  labs(title = "Mean Seasonal Water Salinity by Year")

seasons_sal

```


#Depth

This section will begin by looking at the water depth by season across the years 2002 - 2022.

```{r, echo = FALSE, message = FALSE}

#Selecting relevant data to look at water depth only
depth <- stony_creek |>
  select(Depth, year, season)|>
  filter(!is.na(Depth))

#Creating mean column
mean_depth <- depth |> 
  group_by(season, year)|>
  summarize(mean_depth = mean(Depth))

```

```{r, echo=FALSE, message=FALSE}
#Creating plots by the season
##Spring water depth
depth_spring <- mean_depth |>
  filter(season == "Spring")

line_depth_spring <- ggplot(depth_spring)+
  geom_point(aes(year, mean_depth))+
  labs(title = "Mean Spring Water Depth by Year")

line_depth_spring
```

```{r, echo=FALSE, message=FALSE}
##Fall depth
depth_fall <- mean_depth |>
  filter(season == "Fall")

line_depth_fall <- ggplot(depth_fall)+
  geom_point(aes(year, mean_depth))+
  labs(title = "Mean Fall Water Depth by Year")

line_depth_fall
```

```{r, echo=FALSE, message=FALSE}
##Summer depth
depth_summer <- mean_depth |>
  filter(season == "Summer")

line_depth_summer <- ggplot(depth_summer)+
  geom_point(aes(year, mean_depth))+
  labs(title = "Mean Summer Water Depth by Year")

line_depth_summer
```

```{r, echo=FALSE, message=FALSE}
## Winter depth
depth_winter <- mean_depth |>
  filter(season == "Winter")

line_depth_winter <- ggplot(depth_winter)+
  geom_point(aes(year, mean_depth))+
  labs(title = "Mean Winter Water Depth by Year")

line_depth_winter
```

```{r, echo=FALSE, message=FALSE}
#plotting water depth by all seasons

seasons_depth <- mean_depth|> filter(!is.na(season))|>
  #removed 5 rows
  ggplot()+
  geom_point(aes(year, mean_depth, color = season))+
  scale_color_viridis_d()+
  labs(title = "Mean Seasonal Water Depth by Year")

seasons_depth

```

#Chlorofluorocarbons (ChlFluor)
***ChlFluor starts to be collected at a later date (2014)

This section will begin by looking at the Chlorofluorocarbons levels by season across the years 2002 - 2022.

```{r, echo = FALSE, message = FALSE}

#Selecting relevant data to look at Chlorofluorocarbons only
chloro <- stony_creek |>
  select(ChlFluor, year, season)|>
  filter(!is.na(ChlFluor))

#Creating mean column
mean_chloro <- chloro |> 
  group_by(season, year)|>
  summarize(mean_chloro = mean(ChlFluor))

```

```{r, echo=FALSE, message=FALSE}
#Creating plots by the season
##Spring Chlorofluorocarbon levels
chloro_spring <- mean_chloro|>
  filter(season == "Spring")

line_chloro_spring <- ggplot(chloro_spring)+
  geom_point(aes(year, mean_chloro))+
  labs(title = "Mean Spring Chlorofluorocarbon Levels by Year")

line_chloro_spring
```

```{r, echo=FALSE, message=FALSE}
##Fall Chlorofluorocarbon levels
chloro_fall <- mean_chloro|>
  filter(season == "Fall")

line_chloro_fall <- ggplot(chloro_fall)+
  geom_point(aes(year, mean_chloro))+
  labs(title = "Mean Fall Chlorofluorocarbon Levels by Year")

line_chloro_fall
```

```{r, echo=FALSE, message=FALSE}
##Summer Chlorofluorocarbon levels
chloro_summer <- mean_chloro |>
  filter(season == "Summer")

line_chloro_summer <- ggplot(chloro_summer)+
  geom_point(aes(year, mean_chloro))+
  labs(title = "Mean Summer Chlorofluorocarbon Levels by Year")

line_chloro_summer
```

```{r, echo=FALSE, message=FALSE}
##Winter Chlorofluorocarbon Levels
chloro_winter <- mean_chloro |>
  filter(season == "Winter")

line_chloro_winter <- ggplot(chloro_winter)+
  geom_point(aes(year, mean_chloro))+
  labs(title = "Mean Winter Chlorofluorocarbon Levels by Year")

line_chloro_winter
```

```{r, echo=FALSE, message=FALSE}
#plotting Chlorofluorocarbon Levels by all seasons

seasons_chloro <- mean_chloro|> filter(!is.na(season))|>
  ggplot()+
  geom_point(aes(year, mean_chloro, color = season))+
  scale_color_viridis_d()+
  labs(title = "Mean Seasonal Chlorofluorocarbon Levels by Year")

seasons_chloro

```
```{r, echo=FALSE, warning=FALSE, error=FALSE}
"
Notes: During the meeting on 2/23 we discussed 
Plotting values without separating by season
#Box plots
"
#Plotting mean temp by year

mean_yearly_temperature <- mean_temp|> filter(!is.na(season))|>
  ggplot()+
  geom_boxplot(aes(year, mean_temp))+
  facet_wrap(~year)+
  labs(title = "Mean Yearly Temperature")+
  theme(
    legend.position = "none"
  )

mean_yearly_temperature
```
```{r, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
#Comparing the first and last year of collected data
temp2002Vs2021 <- mean_temp|> filter(!is.na(mean_temp))|>
  filter(year== c(2002, 2021))|>
  ggplot()+
  geom_boxplot(aes(year, mean_temp))+
  facet_wrap(~year)+
  labs(title = "Mean Yearly Temperature")

temp2002Vs2021
```
```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
#Chloro levels by year (not separated by season)
mean_yearly_chloro <- mean_chloro|> filter(!is.na(mean_chloro))|>
  ggplot()+
  geom_boxplot(aes(year, mean_chloro))+
  facet_wrap(~year)+
  labs(title = "Mean Yearly Chlorofluorocarbon Levels")

mean_yearly_chloro
```
```{r, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
#Comparing the first and last year of collected data
chloro2014Vs2021 <- mean_chloro|> filter(!is.na(mean_chloro))|>
  filter(year== c(2014, 2021))|>
  ggplot()+
  geom_boxplot(aes(year, mean_chloro))+
  facet_wrap(~year)+
  labs(title = "Mean Yearly Chlorofluorocarbon Levels")

chloro2014Vs2021
```


