---
title: "variable_exploration"
author: "Joy Terentiev"
format: pdf
editor: visual
---


```{r, echo = FALSE, message=FALSE, warning=FALSE}
#libraries
library(tidyverse)
library(dplyr)
library(lubridate)
library(gt)
library(ggplot2)
library(stringr)
library(janitor)
library(reshape2)
```

Data
```{r}
combined <- read_csv("data/joined_dataset.csv")
```

EDA
```{r}
#overview of variables and NAs
glimpse(combined) 

#summaries of all variables
summary(combined) 
```

Plots

Correlation matrix

```{r}
#select meaningful numeric variables
cor_vars <- combined %>%
  select(
    ATemp, RH, BP, WSpd, MaxWSpd, Wdir, SDWDir, TotPAR, TotPrcp,  #weather
    Temp, DO_mgl, DO_Pct, Sal, SpCond, pH, Turb, ChlFluor, Depth   #water
  ) %>%
  select(where(~sum(!is.na(.)) > 0))#remove NA columns

#compute correlation matrix
cor_matrix <- cor(cor_vars, use = "pairwise.complete.obs")

#remove the upper triangle of the matrix
cor_matrix[upper.tri(cor_matrix)] <- NA

#melt for ggplot
cor_long <- melt(cor_matrix, na.rm = TRUE)
names(cor_long) <- c("Variable1", "Variable2", "Correlation")


#plot lower-triangle matrix
ggplot(cor_long, aes(x = Variable1, y = Variable2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#a6cee3", mid = "#f7f7f7", high = "#fb9a99", midpoint = 0) +
  geom_text(aes(label = round(Correlation, 2)), size = 3) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(title = "Correlation Matrix", fill = "r")
```

Exploring Variables:
1) Depth
2) ChlFl
3) SpCond
4) Temp
5) TotPAR

1. It seems *Depth* has a moderate/weak correlation with:
Temp, DO_mgl, DO_pct, sal, SpCond, ChlFlour, pH, Turb <- WATER
ATemp, BP <- WEATHER

```{r, warning=FALSE, message=FALSE}
#Preparing to plot variable means by season

exploring_depth <- combined|> 
  select(year, season, Temp, DO_mgl,Depth, DO_Pct, Sal, SpCond, ChlFluor,
         ATemp, BP, pH, Turb)|>
  group_by(season,year)|> 
  summarise(mDepth= mean(Depth, na.rm = TRUE),
            mTemp= mean(Temp, na.rm = TRUE),
            mDO_mgl= mean(DO_mgl, na.rm = TRUE),
            mDO_Pct= mean(DO_Pct, na.rm = TRUE),
            mSal= mean(Sal, na.rm = TRUE),
            mSpCond= mean(SpCond, na.rm = TRUE),
            mChlFluor= mean(ChlFluor, na.rm = TRUE),
            mpH= mean(pH, na.rm = TRUE),
            mTurb= mean(Turb, na.rm= TRUE),
            mBP= mean(BP, na.rm = TRUE),
            mATemp= mean(ATemp, na.rm = TRUE))|>
  filter(!is.na(season))
```


Ploting Temp and Depth
```{r}
#Need to fix the color

  ggplot(data= exploring_depth)+
  geom_point(aes(year, mDepth, color="green"))+
  geom_point(aes(year, mTemp, color="blue"))+
  labs(title = "Mean Seasonal Depth and Temperature of Water",
       x= "Year",
       y= "Mean Temp/Depth")+
  theme_minimal()+
  facet_wrap(~season)
```
```{r}
#Need to fix the color

  ggplot(data= exploring_depth)+
  geom_point(aes(year, mDepth, color="green"))+
  geom_point(aes(year, mATemp, color="blue"))+
  labs(title = "Mean Seasonal Depth and Temperature of Air",
       x= "Year",
       y= "Mean Temp/Depth")+
  theme_minimal()+
  facet_wrap(~season)
```

```{r}
  ggplot(data= exploring_depth)+
  geom_point(aes(year, mDepth, color="green"))+
  geom_point(aes(year, mBP, color="blue"))+
  labs(title = "Mean Seasonal Depth and BP",
       x= "Year",
       y= "Mean Temp/BP")+
  theme_minimal()+
  facet_wrap(~season)
```

```{r}
  ggplot(data= exploring_depth)+
  geom_point(aes(year, mDepth, color="green"))+
  geom_point(aes(year, mDO_mgl, color="blue"))+
  labs(title = "Mean Seasonal Depth and DO_mgl",
       x= "Year",
       y= "Mean Depth/DO_mgl")+
  theme_minimal()+
  facet_wrap(~season)
```

```{r}
ggplot(data= exploring_depth)+
  geom_point(aes(year, mDepth, color="green"))+
  geom_point(aes(year, mDO_Pct, color="blue"))+
  labs(title = "Mean Seasonal Depth and DO_pct",
       x= "Year",
       y= "Mean Depth/DO_pct")+
  theme_minimal()+
  facet_wrap(~season)
```

```{r}
ggplot(data= exploring_depth)+
  geom_point(aes(year, mChlFluor, color="green"))+
  geom_point(aes(year, mSal, color="blue"))+
  labs(title = "Mean Seasonal ChlFluor and Sal",
       x= "Year",
       y= "Mean ChlFluor/Sal")+
  theme_minimal()+
  facet_wrap(~season)
```

```{r}
ggplot(data= exploring_depth)+
  geom_point(aes(year, mSpCond, color="green"))+
  geom_point(aes(year, mDO_Pct, color="blue"))+
  labs(title = "Mean Seasonal SpCond and DO_Pct",
       x= "Year",
       y= "Mean SpCond/DO_Pct")+
  theme_minimal()+
  facet_wrap(~season)
```

